\documentclass[11pt, a4paper]{article}

% --- Fonter og tegnsett ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathpazo} % Palatino font for tekst og matte
\usepackage[english]{babel}

% --- Formatering og grafikk ---
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{geometry}
\geometry{a4paper, margin=2.5cm}

% --- Tittel og Forfatter ---
\title{Overlapping Semantic Communities in Word Networks:\\Bridging Partitioning and Polysemy through Line Graph Clustering and Roaring Bitmaps}
\author{Lars G. Bagøien Johnsen \\ Department of Research \\ National Library of Norway \\ lars.johnsen@nb.no}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We introduce a novel approach to resolving the fundamental dilemma between overlapping community detection (k-cliques) and graph partitioning methods (such as the Louvain algorithm) in the context of semantic word networks. While traditional partitioning forces polysemous words into single communities, thereby destroying semantic ambiguity, overlapping methods often suffer from rigidity and low coverage. By applying community detection to the \textit{line graph} of the word network, we demonstrate how edge-based clustering naturally produces overlapping node communities when projected back to the original graph. Furthermore, we address the computational challenge of scaling to corpus-wide co-occurrence data (1.7 billion instances) by utilizing Radon-Nikodym ratios for semantic filtering, encoded into high-performance Roaring Bitmaps. This allows for lightning-fast sub-graph retrieval using purely bitwise logic, circumventing the need for expensive vector-space similarity search.
\end{abstract}

\section{Introduction}
In previous work (Johnsen, 2016), we observed a fundamental trade-off in network semantics when clustering words from collocations. When applying traditional partitioning algorithms like the Louvain method, every word is forced into exactly one community. This approach maximizes global network coverage but completely destroys the representation of polysemy; for instance, the word \textit{kirsebær} (cherry) must arbitrarily be assigned to either the ``fruit'' community or the ``tree'' community. 

Conversely, methods such as k-clique percolation allow nodes to belong to multiple communities simultaneously, successfully capturing the dual meaning of \textit{kirsebær}. However, k-cliques are notoriously strict and conservative, resulting in large portions of the semantic network being left unclustered. The goal of this paper is to bridge these two paradigms, satisfying the need for both high precision overlapping structures and comprehensive graph coverage, a challenge originally identified a decade ago.

\section{Methodology: The Semantic Line Graph}
To solve the partitioning dilemma, we shift the focus of our clustering from the words themselves (the nodes) to the relationships between them (the edges). Inspired by the foundational work on link communities in biological and protein interaction networks (Ahn, Bagrow, and Lehmann, 2010), we transform the original word network $G$ into its corresponding line graph $L(G)$. 

In the line graph, the co-occurrences (edges) from the original graph become nodes. When we apply a highly efficient partitioning algorithm like Louvain to $L(G)$, it cleanly groups the relationships into distinct contextual communities. Because a single word in the original graph $G$ often participates in multiple edges, mapping the partitioned edges back to the original nodes naturally results in words inheriting multiple community memberships. Thus, overlapping polysemy emerges as a mathematical consequence of partitioning the edges.

\subsection{Line Graph Proposals and Centrality Filtering}
We propose treating the community assignments generated by the line graph as semantic ``proposals'' rather than absolute truths. Because language networks are inherently noisy, a word might be connected to a cluster through a single, weak edge (e.g., a rare typographical error or a highly idiosyncratic sentence).

To refine the polysemous groupings, we can apply domain-specific thresholds based on graph topology. By calculating centrality measures—such as Eigenvector Centrality or internal node degree—within the proposed sub-cluster, we can accept or reject a word's membership in a given meaning. This two-step process allows for highly tunable semantic disambiguation, separating robust contextual meanings from spurious statistical artifacts.

\section{Engineering for Scale: Bitwise Semantics}
While line graph clustering provides a robust theoretical solution, constructing semantic networks from corpus-level data introduces immense computational hurdles. In this study, we process over 1.7 billion coordinations extracted from Norwegian newspaper corpora. Expanding nodes to depth $D=3$ in a network of this magnitude leads to an immediate combinatorial explosion, rendering traditional graph traversals unfeasible for real-time applications.

To counteract this, we replace standard floating-point vector similarity with purely bitwise logic. We filter the raw frequencies using the Radon-Nikodym ratio as a measure of semantic ``temperature,'' retaining only connections that exceed strict thresholds (e.g., $RN \ge 100$). The resulting adjacency lists are encoded into Roaring Bitmaps. This compression strategy allows the CPU to perform graph intersections using native bitwise \texttt{AND}/\texttt{OR} operations, achieving sub-millisecond sub-graph retrieval without the use of GPU-accelerated vector databases.

\section{Results and Visualizations}
To demonstrate the efficacy of this architecture, we analyze the semantic neighborhood of highly polysemous words. A primary case study is the Norwegian word \textit{is}, which simultaneously represents frozen water, a dessert (ice cream), and the terrorist organization ISIS. 

Applying line graph clustering to the high-temperature bitmap representations successfully isolates these distinct contexts. The resulting visualizations contrast sharply with earlier k-clique renderings, showcasing perfectly delineated semantic clusters that overlap only at the central, polysemous hubs. Furthermore, by utilizing deterministic random sampling seeded by node IDs, we prevent network explosion at deeper graph layers while ensuring reproducible and mathematically pure visualizations.

\section{Conclusion}
By synthesizing edge-based community detection with bitwise data structures, we have realized the theoretical objective proposed a decade ago. We have demonstrated that the strict partitioning of line graphs provides a comprehensive yet overlapping semantic map of word networks. Furthermore, the use of Roaring Bitmaps ensures that this theoretically rigorous approach remains computationally viable on a multi-billion edge corpus, paving the way for highly scalable semantic web applications.

\begin{thebibliography}{99}
\bibitem{ahn2010}
Ahn, Y. Y., Bagrow, J. P., \& Lehmann, S. (2010). Link communities reveal multiscale complexity in networks. \textit{Nature}, 466(7307), 761-764.

\bibitem{johnsen2016}
Johnsen, L. G. B. (2016). \textit{Graph Analysis of Word Networks}. Department of Research, National Library of Norway.
\end{thebibliography}

\end{document}
